% =============================================================================
% Chapter 2 — Literature Review (REVISED)
% Tara Torbati — Master's Thesis
%
% Revisions in this version:
%   - Sentence-length pass: long compound sentences split into shorter ones
%   - Bridge sentences added at every major section transition
%   - Rhetorical intensifiers stripped throughout
%       ("most recent", "principal", "central", "substantially" etc. removed
%        where they don't carry information)
%   - Tone restored to neutrally analytical (reporting, not arguing)
%   - Em-dash chains and nested qualifications reduced
% =============================================================================

\chapter{Literature Review}
\label{chap:lit}

This chapter surveys the technical literature relevant to constrained
optimal irrigation control on topographically heterogeneous
agricultural fields. Section~\ref{sec:lit-context} situates the
problem within agricultural water management. Sections
\ref{sec:lit-classical} through \ref{sec:lit-rl} review the control
architectures proposed for irrigation, in roughly increasing order of
mathematical sophistication: classical and heuristic controllers,
Model Predictive Control with its modern variants, and Reinforcement
Learning. Section~\ref{sec:lit-spatial} addresses spatially
distributed control and topographic flow modeling, which is central
to the present work but treated only obliquely in most of the
irrigation-control literature. Section~\ref{sec:lit-sensing} reviews
the supporting sensing and communication infrastructure.
Section~\ref{sec:lit-deficit} comments briefly on deficit-irrigation
strategies. Section~\ref{sec:lit-gap} synthesizes the surveyed work
and identifies the specific research gap addressed by this thesis.

\section{Context: Agricultural Water Management Under Scarcity}
\label{sec:lit-context}

Agriculture consumes roughly 70\% of global freshwater
withdrawals~\cite{Sharma2025,Bayar2025}. Demographic pressure is
projected to increase global food demand sharply by mid-century, and
regional water scarcity continues to deepen. Together, these
pressures are pushing the agricultural sector away from
volume-driven heuristic management and toward data-driven precision
systems that match water application to crop demand at finer spatial
and temporal scales than human operators can achieve
manually~\cite{Sharma2025,LopezJimenez2022,Mgana2026}.

This transition is acute in Iran. Surface water and groundwater
resources have been pushed beyond sustainable yield by supply-side
infrastructure expansion, subsidized energy for groundwater pumping,
and weak volumetric controls on agricultural
extraction~\cite{Mesgaran2018,Nouri2023,Behling2022}. In Gilan
Province, the country's principal rice-producing region, heavily
subsidized water tariffs and informal extraction practices have
removed economic constraints on
use~\cite{Mesgaran2018,PourgholamAmiji2024}. The result is a
governance asymmetry in which price signals fail to discipline demand
and conservation measures alone produce systemic rebound effects.
Demand-management approaches in isolation have been argued to be
insufficient~\cite{PerezBlanco2020,Hajirad2025,Hesabi2023}. This
motivates technical controls that operate on the field directly
through water-budget constraints.

Within this technical-control framing, the irrigation problem becomes
a constrained finite-horizon optimal control problem: allocate a
finite volumetric water budget over a fixed agricultural season to
maximize crop yield (or some economic equivalent) subject to per-day
actuator capacities, soil-physical constraints, and crop-physiological
tolerances. The remainder of this chapter reviews the control
architectures proposed to solve this problem.

\section{Classical and Heuristic Controllers}
\label{sec:lit-classical}

The dominant irrigation control regime in deployed agricultural
systems remains heuristic. Fixed schedules calibrated to long-run
climatic norms, threshold-based on/off control triggered by a single
soil moisture sensor, and open-loop time-based timers account for
most installations~\cite{Sharma2025,Bayar2025}. These approaches are
simple to implement, robust to sensor failure, and operationally
familiar. They are appropriate for small-scale or resource-constrained
deployments. They are also fundamentally reactive: they cannot
anticipate impending drought or rainfall, cannot exploit weather
forecasts, and cannot optimize against a global seasonal water
budget. The controller sees only present state, not future demand.

A more sophisticated heuristic class is fuzzy-logic control.
Linguistic rules over input ranges (e.g.\ ``IF soil moisture is LOW
AND temperature is HIGH THEN irrigation is HIGH'') are compiled into
crisp control actions through a defuzzification step. Fuzzy systems
have been integrated with PID feedback to handle nonlinear delayed
dynamics in fertigation systems and canal-gate
control~\cite{Sharma2025,Mahmud2026}. Variants with online
self-correction have been reported to improve transient response in
nonlinear nutrient regulation tasks. Fuzzy-PID architectures address
some limitations of pure PID, but they share with all rule-based
methods the inability to optimize systematically over time. They
also cannot enforce strict resource constraints. A fuzzy controller
has no native mechanism to guarantee that cumulative seasonal water
use stays below a fixed budget.

These limitations matter most in water-scarce contexts where the
binding constraint is not the per-day soil moisture target but the
season-cumulative water allocation. The central question shifts from
``how much water does the crop need today'' to ``how should a finite
seasonal water budget be allocated across 90+ days of varying demand
to maximize terminal yield.'' Classical and heuristic controllers
cannot address this question formally. This motivates the transition
to optimization-based control, taken up in the next section.

\section{Optimal and Predictive Control}
\label{sec:lit-mpc}

\subsection{The Optimal Control Formulation of Irrigation}
\label{ssec:lit-ocp}

The transition from heuristic to optimal control begins with the
formal recognition that irrigation is a constrained finite-horizon
optimal control problem. Following the formulation in
\cite{LopezJimenez2022,Lopezjimenez2024}, the discrete-time problem
is to determine a control sequence
$\mathbf{u}^*(\cdot) = \arg\min_{\mathbf{u}(\cdot)} J(\mathbf{x}(\cdot), \mathbf{u}(\cdot))$
subject to dynamics
$\mathbf{x}(k+1) = f(\mathbf{x}(k), \mathbf{u}(k), \mathbf{w}(k))$,
state constraints (soil moisture remains above the wilting point),
input constraints (per-day actuator capacity), and a season-cumulative
budget constraint
$\sum_k \sum_n u^{(n)}(k) \leq W_{\mathrm{total}}$.
The cost $J$ typically combines a terminal-state reward (terminal
biomass) with running penalties on water consumption, soil moisture
deficit, and surface ponding.

This formulation captures three features that heuristic controllers
cannot. First, an explicit forecast horizon over which future weather
and crop development are anticipated. Second, a global resource
constraint that couples decisions across the entire season. Third, a
multi-objective cost function with weights that admit economic
calibration. The optimization itself is a non-convex nonlinear
program in general, due to crop-physiology nonlinearities and to the
saturation regime of soil drainage. Interior-point or
sequential-quadratic-programming methods are typically required for
solution.

\subsection{Receding-Horizon Implementation: Model Predictive Control}
\label{ssec:lit-mpc-impl}

The standard approach to applying optimal control online is Model
Predictive Control (MPC). At each timestep, MPC solves the optimal
control problem over a finite prediction horizon $H_p$. Only the
first-step control action is applied. The state is then advanced by
one timestep, and the problem is re-solved with updated state and
forecast information. The receding-horizon mechanism transforms an
open-loop optimization into a feedback control law and provides
robustness to modeling errors and
disturbances~\cite{LopezJimenez2022,Sharma2025}.

For irrigation, MPC has been reported to improve water-use efficiency
by 8--29\% relative to heuristic and open-loop schedulers in
greenhouse and open-field studies~\cite{LopezJimenez2022,Mahmud2026}.
The technical challenges are threefold. First, constructing a
control-oriented model that is faithful enough to predict the
hydrological and biological responses while remaining tractable
within the receding-horizon loop. Second, handling the non-smoothness
of crop and soil dynamics (saturation thresholds, $\max$ and $\min$
operators) within smooth-gradient solvers. Third, calibrating the
multi-term cost function to reflect the agronomic and economic
priorities of the application context.

\subsection{Recent Variants and Architectural Extensions}
\label{ssec:lit-mpc-variants}

Several recent variants of irrigation MPC address specific
limitations of the standard formulation. Economic MPC formulations
replace the conventional setpoint-tracking objective with a direct
economic cost that incorporates real-time price signals for water
and energy~\cite{Sharma2025}. Rainfall-Intensity-and-Soil-Properties
MPC (RISPMPC) integrates the Mein--Larson infiltration model to
anticipate runoff and percolation losses under predicted rainfall.
The controller can then deliberately undershoot soil moisture targets
in advance of a storm, so that the soil matrix has capacity to
absorb the rainfall productively. Reported water savings against
conventional MPC reach 30+\% on heavy-clay soils, where infiltration
capacity is the binding constraint~\cite{Sharma2025}.

Distributed MPC architectures address the scalability of the
optimization to large field domains. Negenborn et
al.~\cite{negenborn2009distmpc} formalized distributed MPC for
multi-reach canal networks. The single large optimization is
decomposed into per-subsystem optimizations coupled through
inter-subsystem communication. Subsequent work extended this paradigm
to hierarchical distributed MPC for basin-scale canal networks under
water scarcity, enabling tractable real-time optimization across
irrigation districts spanning thousands of
hectares~\cite{LopezJimenez2022}. These distributed approaches share
with the present work the observation that monolithic MPC scales
poorly with the number of controlled subsystems.

Data-driven MPC variants substitute the physics-based predictive
model with learned surrogates. Long Short-Term Memory networks
trained on offline simulation data have been applied as predictive
models. These approaches trade physical interpretability for
adaptability across heterogeneous field conditions. They have also
been integrated with mixed-integer formulations to handle discrete
actuator levels in spatially variable fields~\cite{Maniam2026}.

The MPC architectures reviewed above all share one practical
limitation: the per-step computational cost of the online
optimization. The next section discusses Reinforcement Learning,
which addresses this limitation by shifting the computational
expense from deployment time to a one-time offline training phase.

\section{Reinforcement Learning for Irrigation}
\label{sec:lit-rl}

\subsection{Why Reinforcement Learning}
\label{ssec:lit-rl-why}

The principal limitation of MPC is the per-step computational burden
of solving the nonlinear program online. Even with warm-starting and
smooth approximations, MPC solve times on field-scale problems range
from seconds at short horizons to minutes or hours at long
horizons~\cite{Lopezjimenez2024}. This places MPC out of reach for
deployment on edge computing devices, low-power embedded controllers,
or systems requiring sub-second control updates.

Reinforcement Learning (RL) addresses this latency limitation
structurally. After offline training in a simulated environment, an
RL policy is a fixed-parameter neural network. Its forward-pass
inference is deterministic and bounded by the network's
fixed-architecture compute requirement: typically sub-millisecond on
commodity hardware regardless of the underlying problem
dimensionality. The training cost is amortized over many millions of
environment interactions but pays off at every deployment step.

\subsection{Applications of RL to Irrigation Scheduling}
\label{ssec:lit-rl-applications}

RL has been applied to irrigation across crop types, geographies,
and algorithm classes. Q-learning agents trained against the
AquaCrop crop-soil simulator have produced policies that outperform
heuristic schedules on lettuce production in Brazil~\cite{Sharma2025}.
Deep RL agents trained with high-dimensional state observations
(soil moisture across multiple depths, leaf-area index, accumulated
rainfall and irrigation, phenological stage) have been reported to
improve profit in open-field wheat production by up to 17\% relative
to conventional heuristic rules~\cite{Bayar2025,Hu2025}. In
controlled-environment agriculture, Advantage Actor-Critic (A2C)
agents have improved water-use efficiency over both threshold-based
and time-based controllers across distinct growth phases of
greenhouse crops~\cite{Bayar2025}.

The practical challenge across these applications is data efficiency.
Direct training on physical fields is infeasible due to the cost of
irreversibly damaging real crops during the exploration phase. The
dominant practice is therefore to train against a high-fidelity
simulator (such as AquaCrop, APSIM, or custom agent-based
models~\cite{Lopezjimenez2024}) and transfer the learned policy to
the physical environment. The fidelity of the simulator becomes the
bottleneck: a poorly-validated simulator produces policies that do
not transfer.

\subsection{Continuous-Action Control: From DDPG to Soft Actor-Critic}
\label{ssec:lit-rl-sac}

The action space for irrigation is continuous: per-agent water
volume in some bounded interval. This rules out value-iteration
methods that operate on discrete action sets. Policy-gradient or
actor-critic methods compatible with continuous controls are
required.

The Deep Deterministic Policy Gradient (DDPG) algorithm of Lillicrap
et al.~\cite{lillicrap2015ddpg} was the first deep RL method to
demonstrate competitive performance on continuous control tasks. It
combines off-policy Q-learning with a deterministic policy gradient
through an actor-critic architecture. DDPG, however, exhibited high
sensitivity to hyperparameters and instability across random seeds.
Twin Delayed Deep Deterministic Policy Gradient
(TD3)~\cite{fujimoto2018td3} addressed several of these failure modes
through twin Q-networks, delayed policy updates, and target-policy
smoothing. TD3 produced more stable training but retained the
deterministic-policy framework.

The Soft Actor-Critic (SAC) algorithm of Haarnoja et
al.~\cite{haarnoja2018sac,haarnoja2019sac} is the current default
for continuous control in deep RL. SAC is built on the
maximum-entropy reinforcement learning framework. The standard
objective $\mathbb{E}_\pi[\sum_t r_t]$ is augmented with an entropy
bonus $\mathcal{H}[\pi(\cdot|s_t)]$ that rewards stochastic
exploration:
\begin{equation}
J(\pi) = \mathbb{E}_\pi\Bigl[\sum_{t=0}^{\infty} \gamma^t
(r_t + \alpha \mathcal{H}[\pi(\cdot \mid s_t)])\Bigr]
\label{eq:sac-objective}
\end{equation}
The temperature parameter $\alpha$ trades off reward maximization
against entropy maximization. It is automatically tuned in the modern
formulation~\cite{haarnoja2019sac} via a constrained Lagrangian update
against a target entropy. The maximum-entropy framework has two
practical benefits relative to deterministic-policy methods. It
produces stochastic policies that explore more effectively during
training. It also produces policies that capture multi-modal optima,
which is useful when several distinct control sequences achieve
similar reward, as occurs in irrigation where the timing of water
application admits considerable latitude. SAC is off-policy (using a
replay buffer) and uses twin Q-networks in the manner of TD3 to
mitigate value overestimation.

For the irrigation problem of this thesis, SAC is adopted as the RL
architecture. The justification is threefold: continuous action
support without quantization artifacts, sample efficiency on the
order required for high-fidelity simulator training, and demonstrated
robustness across hyperparameter and seed variation in the
maximum-entropy formulation~\cite{haarnoja2019sac}.

\subsection{Reward Engineering and Constraint Handling}
\label{ssec:lit-rl-reward}

A central challenge in applying RL to constrained problems is that
standard RL algorithms have no native mechanism for enforcing hard
constraints~\cite{Bayar2025}. The MPC solver, by contrast, treats the
seasonal water budget as a hard linear inequality enforced at every
solver iteration. RL alternatives rely on reward shaping (penalizing
constraint violations) and on action clipping (truncating actions to
the feasible set). These mechanisms shift constraint satisfaction
from a guarantee to an emergent property of training. The reward
function therefore requires careful construction. Documented practice
in irrigation RL combines several elements: a positive baseline
reward for matching crop water demand, an exponential penalty for
over-irrigation that discourages dumping the budget early, a soft
penalty on field-capacity overshoot that prevents waterlogging stress,
and a sparse terminal bonus on harvested biomass that incentivizes
long-horizon yield~\cite{Bayar2025}. The cost function of the
present thesis follows this pattern. It also adopts an explicit
policy-equivalence design choice: the SAC reward is constructed as
the exact negation of the MPC path cost, eliminating the common
confound in MPC-versus-RL comparisons (see Chapter~\ref{chap:controllers}).

\subsection{Multi-Agent and Distributed Reinforcement Learning}
\label{ssec:lit-marl}

When the controlled system consists of many spatially distributed
sub-units, monolithic single-agent RL scales poorly because the
joint action space grows exponentially with the number of agents.
Multi-agent RL (MARL) addresses this by decomposing the problem into
per-agent policies that share information during training but execute
locally. The Centralized Training, Decentralized Execution (CTDE)
paradigm has emerged as the dominant
architecture~\cite{amato2024ctde}. During training, agents have
access to global state and to all other agents' actions through a
centralized critic. At deployment, each agent acts on local
observations only. CTDE methods preserve the coordination benefits
of centralized training while producing policies that are scalable,
communication-free at deployment, and consistent with the partially
observable distributed structure of physical multi-agent systems.

For irrigation on a topographically heterogeneous field, the CTDE
paradigm is naturally applicable. Each agent (corresponding to a
spatial cell of the field) observes its local state. The
training-time critic has access to the joint state and can resolve
the credit-assignment problem of attributing global yield to per-cell
irrigation decisions. The thesis adopts this architecture; the
implementation details are presented in Chapter~\ref{chap:controllers}.

\section{Spatial Modeling and Topographic Heterogeneity}
\label{sec:lit-spatial}

The literature reviewed in Sections~\ref{sec:lit-classical} through
\ref{sec:lit-rl} treats the field, with few exceptions, as a single
lumped soil-moisture reservoir. This simplification is justified for
small, flat, homogeneous plots. It breaks down when the controlled
field is spatially heterogeneous in elevation, soil texture, or crop
distribution. For agricultural fields with non-trivial topography,
including most fields in mountainous regions such as Gilan, lateral
water redistribution through surface runoff couples the moisture
states of adjacent cells in ways that lumped models cannot represent.

\subsection{Variable-Rate Irrigation as a Spatial Control Precedent}
\label{ssec:lit-vri}

The closest commercial precedent for spatially distributed irrigation
control is Variable-Rate Irrigation (VRI). The application rate of a
center-pivot or linear-move sprinkler is modulated as the machine
traverses the field, applying different volumes to different cells
as a function of soil texture, slope, or crop-vigor
maps~\cite{Sharma2025,Bayar2025}. VRI partially addresses the
spatial-heterogeneity problem at the application end. It does not
typically incorporate runoff coupling between cells: each cell's
prescription is computed independently from a static spatial map.
This is a reasonable approximation when slopes are mild and rainfall
events are gentle. It loses fidelity in topographically active
fields where overland flow between cells is non-negligible.

\subsection{DEM-Based Flow Routing}
\label{ssec:lit-flow-routing}

The hydrology literature has developed machinery for representing
surface water flow on a discretized terrain. The foundational
technique is the D8 algorithm of O'Callaghan and
Mark~\cite{ocallaghan1984d8}. Each cell of a digital elevation
model (DEM) is assigned a single downslope flow direction, toward
the neighbor with the steepest descent. The D8 representation
produces a directed graph over the field cells whose edges encode
lateral water exchange. Multiple-flow-direction extensions, including
MD8~\cite{quinn1991mfd}, distribute flow proportionally among all
downslope neighbors weighted by slope. These reduce the artificial
flow-line concentration of D8 at the cost of greater computational
complexity. For control applications, the simpler D8 representation
has been the practical choice in recent agent-based irrigation
work~\cite{Lopezjimenez2024,LopezJimenez2022}, because the routing
graph must be embedded in a symbolic optimization framework with
second-order gradient information.

The integration of DEM-based routing into a control framework
introduces two design considerations not present in standard
hydrology applications. First, the routing must be computable at
every receding-horizon iteration and must be differentiable with
respect to control inputs and state variables. This rules out
iterative routing solvers and discontinuous flow-direction switches.
Second, the routing must respect mass conservation across the field
domain, including at the field boundary. Naive routing schemes
produce degenerate sink behavior in low-elevation cells with no
internal downhill neighbors. The first consideration is addressed by
static topological sorting of the routing graph at NLP construction
time, allowing the forward sweep to be unrolled symbolically. The
second is addressed through fractional routing through padded
boundary cells, as implemented in the present thesis.

\subsection{Agent-Based Models for Spatial Irrigation Control}
\label{ssec:lit-abm}

The line of work that directly addresses spatially distributed
irrigation control on heterogeneous fields is the agent-based model
of Lopez-Jimenez et al.~\cite{Lopezjimenez2024,LopezJimenez2022}.
This work formulates a discrete-cell model in which each agent
maintains its own soil moisture, ponding, and biomass states. Cells
are coupled through D8-derived surface runoff routing. The model is
embedded within a nonlinear MPC formulation for water-budget-
constrained irrigation. The agent-based formulation is compact
relative to fine-resolution finite-element soil hydrology. It admits
the symbolic differentiation required by interior-point optimization,
while preserving the spatial fidelity needed to capture
topographically driven phenomena such as ridge runoff and valley
waterlogging.

The present thesis builds directly on the Lopez-Jimenez framework
with three modifications motivated by the constrained control
setting: an explicit two-layer hydrological decoupling between
surface ponding and root-zone moisture, a padded-DEM boundary
condition that resolves a degenerate sink behavior in the original
formulation, and a coupling of drought stress directly to biomass
accumulation rather than only to maturation. These modifications are
documented in Chapter~\ref{chap:system}.

\section{Sensing, Data Assimilation, and Communication Infrastructure}
\label{sec:lit-sensing}

Field deployment of any closed-loop irrigation controller depends on
the upstream sensing infrastructure that provides the soil moisture,
temperature, and weather observations consumed by the controller.
Soil-based sensing using Time Domain Reflectometry (TDR), Frequency
Domain Reflectometry (FDR), and capacitive probes is the dominant
modality, deployed at cell-level density across the
field~\cite{Sharma2025,Bayar2025}. Plant-based sensing of leaf
turgor, sap flow, and stomatal conductance is increasingly used in
high-value horticulture for direct measurement of physiological water
status. Atmospheric sensing using satellite remote sensing products
(SMAP, SMOS, MERRA-2 GWETROOT~\cite{Afrasiabikia2026}) provides
field-mean estimates that complement localized in-situ measurements
and are useful for validation of integrated field-scale models.

The practical challenges are sensor drift in harsh field environments
and the energy and bandwidth constraints of farm-wide deployment.
Sensor drift motivates online data assimilation through particle-
filter or Kalman-filter calibration~\cite{Maniam2026}. The
energy-bandwidth constraint motivates Low-Power Wide-Area Network
(LPWAN) protocols such as LoRa and NB-IoT for telemetry. These
infrastructure questions are critical for deployment but are largely
orthogonal to the control-policy question of this thesis. They are
not developed further.

\section{Deficit Irrigation and Crop-Physiological Constraints}
\label{sec:lit-deficit}

A substantial literature addresses deficit-irrigation strategies in
which water application is deliberately reduced below crop demand
during non-critical phenological stages. The plant's physiological
response to mild water stress is exploited to improve fruit quality
and water-use efficiency. Regulated Deficit Irrigation (RDI) and
Partial Root-zone Drying (PRD) have been reported to achieve
substantial water savings in stone fruits, processing tomatoes, and
citrus orchards while preserving or improving harvested
quality~\cite{Sharma2025}. These strategies are compelling for crops
in which water stress modulates harvest quality through
secondary-metabolite accumulation. They are largely inapplicable to
flooded-paddy rice cultivation. Rice's intolerance of moderate-to-
severe drought stress~\cite{jalali2021,raoufi2025} constrains the
controller to keeping soil moisture above the agronomic stress
threshold throughout the season. The present thesis operates in this
constrained regime and does not pursue deficit-irrigation strategies.
The deficit budget scenarios at 85\% and 70\% allocation evaluated in
Chapter~\ref{chap:results} test controller robustness under hard
volumetric scarcity, not under deliberate stress imposition.

\section{Synthesis and Research Gap}
\label{sec:lit-gap}

The literature surveyed in this chapter establishes the principal
families of control architectures applied to the irrigation problem
and their respective strengths and limitations. Heuristic and
classical controllers are simple and operationally robust but cannot
enforce season-cumulative constraints or anticipate future weather.
MPC provides constraint-aware, forecast-driven optimization but at a
per-step computational cost that limits edge-device deployment and
that scales unfavorably with prediction horizon. RL provides
sub-millisecond inference after training but lacks native constraint
guarantees and depends critically on the fidelity of the training
simulator. Each family has been demonstrated in irrigation
applications, and recent work has produced increasingly sophisticated
extensions: RISPMPC, distributed MPC, CTDE-based multi-agent RL.

Within this landscape, three gaps motivate the present thesis. First,
the integration of agent-based topographic cascade routing with
constrained MPC has been demonstrated in only a handful of recent
works~\cite{Lopezjimenez2024,LopezJimenez2022}. Several pathologies
of the resulting NLP --- including degenerate sink behavior,
field-capacity overshoot at long horizons, and chronic waterlogging
in wet-year scenarios --- have not been systematically characterized
or resolved. Second, the calibration of the multi-term MPC cost
function to real economic price tiers has not been demonstrated in
the published literature. Existing work treats the weights as
abstract regularization constants rather than as quantities anchored
to volumetric water tariffs. Third, the direct head-to-head
comparison of constrained MPC against a SAC controller trained on
the same plant model and against an exact-equivalent reward function
has not been reported for spatially distributed agricultural
control. Such a comparison enables rigorous policy comparison rather
than algorithm-versus-algorithm benchmarking.

The thesis addresses these three gaps. Chapter~\ref{chap:system}
presents the validated agent-based virtual plant on which all
controllers operate. Chapter~\ref{chap:controllers} introduces a
constrained MPC formulation with a six-term cost function (later
reduced to five terms via factorial sensitivity analysis) anchored
to the four-tier Iranian water pricing schedule, and a SAC controller
trained against the exact negation of the MPC path cost.
Chapter~\ref{chap:results} reports the comparative performance of
both controllers across nine combinations of climate scenario and
budget allocation. The Iranian context provides both the policy
motivation, through the water-scarcity dynamics surveyed in
Section~\ref{sec:lit-context}, and the empirical anchor, through
the Gilan rice-production setting on which the virtual plant is
parameterized.
